\chapter{Interrupts}
\label{ch:interrupts}

The context switch in the previous chapter does not run by magic -- something
has to \emph{call} the scheduler. On this port that something is almost always an
interrupt: the level-5 timer tick, or a device interrupt whose handler makes a
higher-priority task ready. This chapter follows an interrupt from the silicon
vector all the way to an Ada handler, and explains how \texttt{pragma
Attach\_Handler} is supported across all three profiles -- including the
bare-board full \texttt{System.Interrupts} the \texttt{full} profile needs,
which no stock bare-board runtime provides.

\section{The Xtensa interrupt levels}\index{interrupts}\index{interrupt levels}\index{xt\_highint5}

The ESP32-S3's Xtensa LX7 has six maskable interrupt levels plus a non-maskable
one, and two of those levels are special:

\begin{longtable}{p{2.2cm} p{10.5cm}}
\toprule
\textbf{Level} & \textbf{Role on this port} \\ \midrule
\endhead
1     & The general \emph{exception} vector (synchronous faults). No async
        interrupt is assigned here. \\
2--3  & Device interrupts (\texttt{\_\_gnat\_level2\_vector} /
        \texttt{\_\_gnat\_level3\_vector}). \\
4     & \emph{Unused} -- there is no level-4 vector; \texttt{EXCSAVE\_4} is
        borrowed as a scratch register by the level-5 vector. \\
5     & \textbf{Kernel-reserved}: the CCOMPARE2 timer tick (interrupt 16) and the
        cross-core IPI ``poke'' (interrupt 31). \\
6     & \texttt{DEBUGLEVEL} -- debug exceptions: breakpoints and the
        \texttt{DBREAK} data watchpoints (used for stack-overflow detection). \\
7     & \texttt{NMILEVEL} -- non-maskable; not used. \\
\bottomrule
\end{longtable}

Two architectural constants shape everything below. \texttt{XCHAL\_EXCM\_LEVEL}
is 3, so \texttt{PS.EXCM} masks only levels 1--3; levels 4--6 are ``high-level''
interrupts with their own dedicated vectors and their own \texttt{EPC$n$/EPS$n$}
save registers, and they keep firing even while a low-level exception is being
taken. And the kernel owns level 5 exclusively -- no application interrupt
priority may mask it, or it would silence the scheduler's own tick (this is why
the Ada interrupt-priority ceilings are capped one below it; see the context-switch
and CXD1006 discussions).

\section{Priorities and the masking model}\index{interrupt priorities}\index{interrupt masking}

Ada interrupt priorities map onto Xtensa levels in
\texttt{System.BB.Board\_Support.Priority\_Of\_Interrupt}: a hardware interrupt's
fixed level becomes the Ada priority its handler runs at, and conversely a task's
ceiling becomes the \texttt{INTLEVEL} it raises while it holds a lock
(\texttt{Enable\_Interrupts} in \texttt{s-bbcppr}). \texttt{Ada.Interrupts.Names}
gives symbolic IDs for the slots that are free for application use:
\texttt{Device\_L2\_0..2} (CPU interrupts 19/20/21), \texttt{Device\_L3\_0/1}
(23/27), with ceilings \texttt{Device\_L2\_Priority} / \texttt{Device\_L3\_Priority}.
The tick (16) and poke (31) are listed as kernel-reserved -- do not attach to
them.

\section{From silicon to handler}\index{interrupt vector}\index{interrupt wrapper}

\texttt{start.S} points \texttt{VECBASE} at our vendored vector table before
anything else runs, so every vector is ours. When a level-5 interrupt fires
(say the tick), this is the path:

\begin{enumerate}
  \item \textbf{Vector entry} (\texttt{xt\_highint5}, \texttt{highint5.S}). The
        CPU jumps to the level-5 slot. The handler first checks one thing
        \emph{without touching the stack}: was the interruptee inside a window
        over/underflow handler (\texttt{EPC\_5} in the window-vector range)? If so
        its \texttt{a1}/SP is transiently invalid, so the tick is \emph{deferred}
        -- push CCOMPARE2 forward and \texttt{rfi} back, letting the window
        handler finish (the CXD8002 fix).
  \item \textbf{Save.} Otherwise it builds an \texttt{XT\_STK} frame on the
        interrupted task's stack and calls \texttt{\_xt\_context\_save} to spill
        the full register state into it. (Debug is masked across this frame build
        so the spill cannot trip the per-task stack-overflow watchpoint -- see
        the context-switch chapter.)
  \item \textbf{Service.} It drops to \texttt{INTLEVEL 5} and calls the Ada
        dispatch routine -- \texttt{\_\_gnat\_timer\_interrupt} for level 5,
        \texttt{\_\_gnat\_level3\_dispatch} / \texttt{\_\_gnat\_level2\_dispatch}
        for the device vectors. Each reads the Xtensa \texttt{INTERRUPT} register,
        and for a pending source it owns it calls
        \texttt{System.BB.Interrupts.Interrupt\_Wrapper (Id)}.
  \item \textbf{The wrapper} (\texttt{s-bbinte.adb}) is the GNARL bridge: it
        records the interrupt being handled, raises the running task's active
        priority to the interrupt's priority (so the ceiling protocol holds and
        lower interrupts stay masked), runs the \emph{registered handler}, then
        restores priority. The handler is an
        \texttt{access procedure (Id : Interrupt\_ID)}.
  \item \textbf{Epilogue -- the deferred switch.} Back in the vector, if the
        dispatch made a higher-priority task ready (\texttt{Context\_Switch\_-
        Needed}) the switch is performed \emph{here}, from the vector's clean
        single-window context, by tail-jumping \texttt{\_\_gnat\_preempt\_-
        dispatch}. The kernel never runs the cooperative spill from interrupt
        context -- it sets a \texttt{Switch\_Pending} flag and lets the epilogue
        do it (Option~A; the previous chapter).
  \item \textbf{Resume.} If no switch is due, \texttt{\_xt\_context\_restore} +
        \texttt{rfi 5} returns to the interruptee atomically (INTLEVEL stays high
        until the very last instruction, closing the re-entrancy window).
\end{enumerate}

A device interrupt at level 2 or 3 follows the identical shape one level down
(\texttt{EPS\_3/EPC\_3}, \texttt{rfi 3}, etc.), and may itself be preempted by the
level-5 tick.

\section{Attaching a handler: two axes}\index{Attach\_Handler}\index{protected handler}

It helps to keep two \emph{independent} distinctions apart. The first is
\emph{which API layer} you attach through; the second -- which lives entirely
inside the high-level layer -- is \emph{static versus dynamic} attachment.

\subsection{Axis 1 -- API layer (high level vs low level)}\index{Ada.Interrupts}

\textbf{High level -- \texttt{Ada.Interrupts} / \texttt{pragma Attach\_Handler}.}
The portable Ada way: the handler is a protected procedure, the enclosing
protected object's \texttt{Interrupt\_Priority} is its ceiling, and the compiler
generates the registration and ceiling locking. This is what the GPIO HAL and the
\texttt{esp32s3\_intr\_levels} example use. The same counting handler as the
low-level version below, written this way, is:

\begin{lstlisting}
with Ada.Interrupts.Names; use Ada.Interrupts.Names;

package body High is
   protected PO with Interrupt_Priority => Device_L2_Priority is
      function Count return Natural;
   private
      procedure Handle;
      pragma Attach_Handler (Handle, Device_L2_0);  -- installed at elaboration
      N : Natural := 0;
   end PO;

   protected body PO is
      procedure Handle is           -- runs at the PO ceiling; locking is implicit
      begin
         Clear_Source;              -- ack the (level-triggered) source
         N := N + 1;
      end Handle;
      function Count return Natural is (N);
   end PO;
end High;
\end{lstlisting}

There is no \texttt{Install} call: \texttt{pragma Attach\_Handler} wires the
handler up when \texttt{PO} elaborates. The handler is a protected procedure (no
\texttt{Id} parameter -- it is bound to one interrupt), the count \texttt{N} is a
protected component rather than a bare \texttt{Volatile}, and the ceiling protocol
serialises it against other accesses to \texttt{PO} for free. Contrast the
hand-rolled low-level form:

\textbf{Low level -- \texttt{System.BB.Interrupts.Attach\_Handler}.} The
bareboard kernel's own primitive: register a plain \texttt{access procedure (Id)}
for an \texttt{Interrupt\_ID}, with an explicit priority. This is what
\texttt{Interrupt\_Wrapper} ultimately calls, it is available on \emph{all}
profiles, and it needs no protected object -- but also gives you no
compiler-enforced ceiling. It is the escape hatch when the portable layer is
unavailable or unwanted:

\begin{lstlisting}
with System; use System;
with System.BB.Interrupts; use System.BB.Interrupts;

package body Low is
   Count : Natural := 0 with Volatile;

   procedure Handle (Id : Interrupt_ID) is
   begin
      Clear_Source (Id);          --  ack the (level-triggered) source
      Count := Count + 1;
   end Handle;

   procedure Install is
   begin
      --  plain procedure, explicit Id + priority, no protected object
      Attach_Handler (Handle'Access, Id => 19,
                      Prio => Interrupt_Priority'Last);
   end Install;
end Low;
\end{lstlisting}

\subsection{Axis 2 -- static vs dynamic (within the high-level layer)}\index{dynamic attachment}

Static versus dynamic is the dividing line of the \texttt{No\_Dynamic\_Attachment}
restriction; it only concerns the high-level \texttt{Ada.Interrupts} layer.

\textbf{Static -- \texttt{pragma Attach\_Handler}.} The handler is fixed at
compile time and installed once, automatically, when the protected object is
elaborated. There is no run-time attach call; the compiler lowers it straight to
\texttt{System.Interrupts.Install\_Handlers}, \emph{bypassing}
\texttt{Ada.Interrupts}. This is the form to reach for by default:

\begin{lstlisting}
with Ada.Interrupts.Names; use Ada.Interrupts.Names;

package body Blink is
   protected L2_PO with Interrupt_Priority => Device_L2_Priority is
   private
      procedure Handle;
      pragma Attach_Handler (Handle, Device_L2_0);  --  installed at elaboration
      N : Integer := 0;
   end L2_PO;

   protected body L2_PO is
      procedure Handle is
      begin
         Clear_L2;                 --  ack the source
         N := N + 1;
      end Handle;
   end L2_PO;
end Blink;
\end{lstlisting}

\textbf{Dynamic -- the \texttt{Ada.Interrupts} operations.} Here the handler is
attached, swapped, or detached \emph{at run time}. The protected procedure must
first be made eligible with \texttt{pragma Interrupt\_Handler}; then
\texttt{Attach\_Handler}, \texttt{Detach\_Handler}, \texttt{Exchange\_Handler} and
\texttt{Current\_Handler} act on it during execution. These calls go \emph{through}
the \texttt{Ada.Interrupts} body -- the path the bare runtime initially stubbed
out (every operation raised \texttt{Program\_Error}) until a delegating
\texttt{a-interr} body was supplied (see \emph{Why the profiles differ} below):

\begin{lstlisting}
with Ada.Interrupts;       use Ada.Interrupts;
with Ada.Interrupts.Names; use Ada.Interrupts.Names;

package body Dyn is
   protected PO is
      procedure Handle;
      pragma Interrupt_Handler (Handle);   --  eligible for dynamic attach
   end PO;
   protected body PO is
      procedure Handle is begin null; end Handle;
   end PO;

   procedure Run is
   begin
      Attach_Handler (PO.Handle'Access, Device_L2_0);  --  attach now
      --  ... later, restore the default treatment ...
      Detach_Handler (Device_L2_0);
   end Run;
end Dyn;
\end{lstlisting}

So the two axes compose: a high-level handler is either static (the \texttt{pragma},
installed at elaboration) or dynamic (the run-time \texttt{Ada.Interrupts} calls),
while the low-level \texttt{System.BB.Interrupts} primitive sits beneath both --
itself a run-time call, but \emph{not} the Ada \texttt{No\_Dynamic\_Attachment}
notion of ``dynamic''.

\section{Why the profiles differ}\index{runtime profile}\index{System.Interrupts}

The high-level form compiles differently depending on the tasking profile, and
that is the crux:

\begin{itemize}
  \item \texttt{embedded} and \texttt{light-tasking} set \texttt{pragma Profile
        (Jorvik)}. Under a Ravenscar-family profile the compiler lowers
        \texttt{pragma Attach\_Handler} to the \emph{restricted} installation path,
        \texttt{System.Interrupts.Install\_Restricted\_Handlers} -- which the
        bareboard \texttt{s-interr} provides. So it works.
  \item \texttt{full} sets \emph{no} profile pragma (its whole point is to lift
        the Jorvik restrictions). The compiler then lowers interrupt-handler
        protected objects to the \emph{full dynamic} machinery --
        \texttt{Register\_Interrupt\_Handler}, \texttt{Install\_Handlers}, and the
        \texttt{Static\_Interrupt\_Protection} / \texttt{Dynamic\_Interrupt\_-
        Protection} protected types.
\end{itemize}

(Note this is the \emph{profile} distinction, not \texttt{Configurable\_Run\_Time},
which is \texttt{True} in both.) The catch is that AdaCore ships a full
\texttt{System.Interrupts} only for hosted (POSIX) targets, where it is built on
a signal-server task and is unportable to bare metal; \emph{no} bare-board
runtime -- not even AdaCore's own \texttt{ravenscar-full} -- implements the
full surface because Ravenscar forbids dynamic attachment in the first place.
For a long time, then, a \texttt{full}-profile build of an \texttt{Ada.Interrupts}
handler simply failed with \texttt{construct not allowed in this configuration}
and \texttt{System.Interrupts.Register\_Interrupt\_Handler not defined}.

\subsection{The bare-board full \texttt{System.Interrupts}}\index{umbrella handler}

That gap is now closed. \texttt{full\_overlay/gnarl/s-interr}
is a bare-board re-implementation of the full \texttt{System.Interrupts}
surface -- \texttt{Register\_Interrupt\_Handler}, the \texttt{Static\_-} and
\texttt{Dynamic\_Interrupt\_Protection} protected bases, \texttt{Install\_-
Handlers} and the dynamic \texttt{Attach\_}/\texttt{Detach\_}/\texttt{Exchange\_-
Handler} operations -- but layered on the machinery that already exists rather
than on a signal server. It only had to add the GNARL-facing surface; the hard
parts were already done and proven (the previous sections):

\begin{itemize}
  \item It installs a single \emph{umbrella} handler per interrupt through the
        low-level \texttt{System.OS\_Interface.Attach\_Handler}
        (= \texttt{System.BB.Interrupts}), then redirects it by re-pointing a
        handler table. The bare-board primitive accepts only \emph{one} attach
        per source, so the umbrella is installed once and the table does the rest.
  \item The interrupt's run priority is fixed by its Xtensa level, so the
        controller-programming step ignores the priority argument
        (\texttt{Install\_Interrupt\_Handler} marks it \texttt{Unreferenced}).
        That is what makes the dynamic \texttt{Attach\_Handler} path -- which
        carries no priority -- safe.
  \item Interrupt \emph{entries} and the POSIX.5 signal services have no
        bare-board meaning and raise \texttt{Program\_Error}.
\end{itemize}

The overlay file replaces the restricted \texttt{s-interr} for the \texttt{full}
profile \emph{only}; \texttt{embedded} and \texttt{light-tasking} keep the
restricted one they need for the Jorvik path. So the same protected-handler
source now builds on all three profiles.

There is one wrinkle, and it is subtler than it first looks. The full
\texttt{System.Interrupts} alone is necessary but not \emph{sufficient}: the
\emph{static} \texttt{pragma Attach\_Handler} path reaches it directly (the
compiler lowers the handler to \texttt{Install\_Handlers}), but the \emph{dynamic}
operations go through \texttt{Ada.Interrupts}, and the bare runtime inherited the
\emph{restricted} (\texttt{No\_Dynamic\_Attachment}) \texttt{Ada.Interrupts} body
from the Jorvik base -- in which every operation is simply \texttt{raise
Program\_Error}. So for a while static attach worked while
\texttt{Ada.Interrupts.Attach\_Handler}, \texttt{Detach\_Handler},
\texttt{Exchange\_Handler}, \texttt{Current\_Handler}, \texttt{Is\_Attached} and
\texttt{Reference} all raised \texttt{Program\_Error} (the full ACATS sweep caught
this -- \texttt{CXC3007}). The overlay now also supplies the \emph{delegating}
\texttt{a-interr.adb} that forwards each operation to \texttt{System.Interrupts},
so the full dynamic API works (\texttt{CXC3007} HW-verified \texttt{PASSED}).

The genuine residual limitation is narrow: because \texttt{Configurable\_Run\_Time}
is \texttt{True}, the front end does not generate library-level finalization, so a
\emph{library-level} interrupt protected object's \texttt{Finalize} -- which would
restore the previously installed handlers at program shutdown -- does not run
automatically. Explicit \texttt{Detach\_Handler} works; only the implicit
restore-on-finalization of a library-level handler PO is absent. That is harmless
on an embedded target that never exits, and the low-level
\texttt{System.BB.Interrupts.Attach\_Handler} primitive remains available on all
profiles as the no-finalization escape hatch.

\section{Exercising the vectors}\index{CPU\_INT}

Because a bug in the vectors (a corrupted context, a dropped interrupt) is exactly
the kind of thing that hides until the worst moment, the
\texttt{esp32s3\_intr\_levels} example is a standalone regression test for them. It
fires the level-2 and level-3 device vectors with \emph{no external wiring} by
routing the \texttt{FROM\_CPU} interrupt-matrix sources (the same mechanism as the
cross-core poke) to \texttt{Device\_L2\_0} and \texttt{Device\_L3\_0} and asserting
them from software, while a low-priority victim task holds floating-point
accumulators and a \texttt{THREADPTR} sentinel live in registers -- continually
preempted by those two vectors \emph{and} the level-5 tick. If any vector failed
to preserve the interrupted context, an accumulator or the sentinel comes back
wrong. On hardware the handler counts climb and the corruption count stays zero,
confirming all three vectors both dispatch correctly and preserve context.
(Level~4 has no vector to test; level~1 carries no asynchronous interrupt on this
port.)

The companion \texttt{esp32s3\_full\_intr} example is the same test built under
the \texttt{full} profile. Its job is narrower: to prove that the identical
\texttt{pragma Attach\_Handler} protected objects now build, bind, link \emph{and
run} against the bare-board full \texttt{System.Interrupts} described above --
the exact source that used to fail with \texttt{Register\_Interrupt\_Handler not
defined} on \texttt{full}. On hardware it boots clean on both cores and the L2
and L3 handler counts climb together with the clean-batch counter, with zero
context-loss and zero faults -- so on \texttt{full}, too, the portable
\texttt{Ada.Interrupts} protected-handler form dispatches correctly and preserves
the preempted context.
