\chapter{Formal Proof with SPARK}
\label{ch:spark}\index{SPARK}\index{formal proof}\index{proof!SPARK}

Every layer in Chapter~\ref{ch:testing} answers the same question in a different
way: \emph{does the code do the right thing on the inputs we tried?} A test runs one
input and checks one answer; a thousand tests run a thousand. Formal proof asks a
different question, and answers it once for all inputs: \emph{can this code go wrong
on \textbf{any} input at all?} Where a test demonstrates the presence of correct
behaviour on an example, a proof demonstrates the \emph{absence} of a whole class of
failure over the entire input space. That is the layer this chapter adds, and for a
bare-metal system that parses bytes it never chose --- a certificate from a stranger,
a GPS sentence from a noisy sky, a DNS reply from the network --- it is the layer
that matters most.

SPARK is a subset of Ada with a proof tool, \texttt{gnatprove}, that ships in the
same Alire toolchain as the compiler. Mark a package \lstinline{with SPARK_Mode => On}
and the tool reads its contracts and its code and discharges, or fails to discharge,
a \emph{verification condition} for every operation that could fault: every array
index, every arithmetic operation that could overflow, every division, every
dereference, and the termination of every loop. When it succeeds it has proved, with
an SMT solver rather than a test, that the code is free of run-time errors --- the
\emph{silver} level of SPARK adoption, absence of run-time errors (\index{AoRTE}AoRTE).
No \lstinline{Constraint_Error}, no \lstinline{Storage_Error} from an unbounded loop,
no buffer overrun, on any input the types allow.

\section{What is proved}\index{proof!surface}

The proof surface is the pure, bounded logic of the HAL: the parsers, serializers,
checksums, and conversions that compute over bytes rather than poke registers. Around
thirty units carry \lstinline{SPARK_Mode => On} and discharge some seven hundred and
eighty run-time checks with zero unproved. The set falls into natural groups: the
ext4 on-disk serializers and their byte helpers (Chapter~\ref{ch:storage}); the
network and protocol parsers; the software half of the cryptography; and a scattering
of pure arithmetic --- civil-date math, longest-prefix routing, sensor conversions,
and the timing dividers extracted from the PWM drivers.

The headline is the group exposed to hostile input. The X.509 certificate reader, the
NMEA GPS-sentence parser, the DNS response parser, the certificate-chain walk, and the
Modbus slave and master all parse bytes chosen by someone else, and all are now proved
to have \textbf{no buffer overrun, no overflow, and no infinite loop on any malformed
or malicious message}. This is not a hardened version that passed a fuzzer; it is a
proof that the failure cannot occur. The DNS parser makes the point sharply: a DNS
name can contain a \emph{compression pointer} back into the message, and a malicious
pointer that points at itself would loop a naive parser forever. The proof required a
\lstinline{Loop_Variant} showing the cursor strictly advances, so termination on a
self-referential pointer is not tested but guaranteed.

\section{Drawing the boundary}\index{SPARK!boundary}

A driver is not pure. It writes volatile registers, follows access-type pointers, and
raises exceptions on bad hardware --- none of which is in the SPARK subset. The
technique is to draw a line \emph{inside} the unit. The pure computation is marked
\lstinline{On} and proved; the parts that touch the world are marked
\lstinline{SPARK_Mode => Off} and left to the other verification layers. For the ext4
serializers this meant \emph{factoring}: the buffer-to-record step was lifted out of
each \lstinline{Read}/\lstinline{Write} into a pure \lstinline{Decode}/\lstinline{Encode}
function that a proved caller invokes, while the block-device I/O and the checksum
stayed behind the boundary. The move is behaviour-neutral --- the same expressions,
relocated --- and the host filesystem suite (Chapter~\ref{ch:testing}, Layer~2) reran
\texttt{e2fsck}-clean afterward to confirm it.

One recurring obstacle is worth recording because it shapes every buffer proof. An
\lstinline{array (Natural range <>)} of bytes can, \emph{in the type}, have a length
of $2^{31}$, which overflows \lstinline{Integer}; so the solver cannot even prove that
\lstinline{Off + N <= Buf'Length} does not overflow, and the whole parser stalls. The
fix is to constrain the index --- \lstinline{subtype Buffer_Index is Natural range 0 .. 2**24 - 1}
--- so the length provably fits an integer. A certificate or a filesystem block is far
below sixteen megabytes, so the constraint costs nothing and unlocks everything; it is
the single change that made the ext4 and X.509 buffers provable.

\section{What the proofs found}\index{bug!found by proof}

A proof that goes through is quiet. A proof that will not go through is often pointing
at a real defect, and this effort surfaced five, every one on a malformed-input path
that testing had not reached:

\begin{itemize}
  \item \textbf{Modbus \lstinline{Put_MBAP}} computed \lstinline{PDU_Len + 1} on an
    unbounded length --- an overflow on a crafted request. The proof drove the
    precondition \lstinline{PDU_Len <= Max_PDU}, the real 253-byte protocol cap.
  \item \textbf{Superblock \lstinline{Encode}} checksummed an \emph{absolute}
    \lstinline{Buf (Base .. Base + N)} slice where every writer used
    \lstinline{Buf'First + offset} --- wrong for a non-zero-based buffer; now
    \lstinline{Buf'First}-relative.
  \item \textbf{X.509 \lstinline{Host_Matches}} could index out of range on a hostile
    certificate whose subject-alt-name slices did not match its buffer; now guarded.
  \item \textbf{Chain validation} dereferenced a certificate's \lstinline{Data} pointer
    with no null or empty check --- a caller-supplied chain could crash it; now rejected
    as malformed.
  \item \textbf{The SHT41 CRC} loop bound \lstinline{while I + 2 <= Data'Last} overflowed
    in its own guard and underflowed on an empty buffer; rewritten as a group count.
\end{itemize}

None was reachable by a well-formed message, which is exactly why the test suites had
not caught them and why proving over the whole input space did.

\section{The edge of the subset}\index{SPARK!limits}

Proof has an honest boundary, and naming it is part of the discipline. Three things sit
outside it. First, code that is \emph{out of the SPARK subset by construction}: the
volatile memory-mapped register drivers, the hardware crypto accelerators
(SHA, AES, RSA), the controlled-type bus sessions with their \lstinline{Finalize}, the
access-to-subprogram interrupt callbacks. These are verified by the other layers ---
loopback self-tests and known-answer tests --- not by proof, and that is the right tool
for them. Second, pure arithmetic that is \emph{inlined} in a register procedure, like a
baud prescaler or a dead-time divider; that logic \emph{can} be proved, but only after
extracting it into a pure sibling package, which is exactly what was done for the
TWAI, LEDC, RMT, and MCPWM timing math. Third, code that is provable \emph{in principle
but not in practice}: the P-256 signature check is legal SPARK and the tool runs, but
proving it pulls in the entire vendored SPARKNaCl elliptic-curve contract closure, which
did not converge under a long budget --- a dedicated, lemma-level effort rather than a
harvest. Knowing which of the three a unit falls into is what keeps the surface honest.

\section{Running the proofs}\index{book!prove}

The proofs run the way the tests do, from one script. \texttt{book/prove/prove.sh}
invokes \texttt{gnatprove} at the silver level over each native prove project and exits
non-zero if any run-time check is unproved, so a change that breaks a proof breaks the
build just as a failing test would. The proofs run host-native because run-time-error
freedom of pure logic is target-independent --- the same argument as the host tests,
one layer up. The companion \texttt{book/prove/README.md} lists the proved units, the
factoring pattern, and the toolchain gotchas learned along the way. Formal proof does
not replace the layers below it; it caps them. The compiler rejects the ill-typed, the
host tests check the answers, the self-tests check the wiring, the sweep checks the
foundation --- and above them all, proof rules out an entire category of failure on
inputs no test will ever enumerate.
