\chapter{Mass Storage: SD Cards, SPI Flash, and the ext4 Filesystem}
\label{ch:storage}

The final layer is mass storage: block drivers for SD cards and for a raw SPI NOR
flash chip, and a complete pure-Ada ext2/3/4 filesystem (a reimplementation of
lwext4) that sits on top of them --- mounting real \texttt{mkfs.ext4} cards,
\emph{and} wear-leveling and formatting a bare flash chip entirely on-device.

\section{SD\_SPI --- an SD card over the SPI master}\index{SD card}\index{SD\_SPI}\index{SPI SD mode}

\texttt{ESP32S3.SD\_SPI} talks to an SD/SDHC card in SPI mode, layered on the
task-safe SPI driver. The chip-select is an ordinary GPIO so it can be held across
a whole command. Blocks are 512-byte logical sectors (LBA addressing); SDSC and
SDHC are detected and handled internally.

\begin{lstlisting}
type Block is array (0 .. 511) of Interfaces.Unsigned_8;
type Status is (OK, No_Card, Unusable, Init_Timeout, Read_Error, Write_Error);
procedure Setup (C : out Card; Host : ESP32S3.SPI.SPI_Host;
                 Sclk, Mosi, Miso, Cs : Pin_Id;
                 Init_Clock_Hz : Positive := 400_000;
                 Data_Clock_Hz : Positive := 8_000_000);
procedure Initialize (C : in out Card; Result : out Status);
procedure Read_Block  (C : in out Card; LBA : Block_Address;
                       Data : out Block; Result : out Status);
procedure Write_Block (C : in out Card; LBA : Block_Address;
                       Data : Block; Result : out Status);
\end{lstlisting}

\textbf{Example 1 --- bring up a card and read sector 0.}
\begin{lstlisting}
with ESP32S3.SD_SPI; use ESP32S3.SD_SPI;
with ESP32S3.SPI;
procedure SD_Init is
   C   : Card;
   St  : Status;
   Sec : Block;
begin
   Setup (C, ESP32S3.SPI.SPI2, Sclk => 12, Mosi => 11, Miso => 13, Cs => 10);
   Initialize (C, St);
   if St = OK then
      Read_Block (C, LBA => 0, Data => Sec, Result => St);   --  the MBR sector
   end if;
end SD_Init;
\end{lstlisting}

\textbf{Example 2 --- a non-destructive read / write-back / re-read round-trip.}
\begin{lstlisting}
with ESP32S3.SD_SPI; use ESP32S3.SD_SPI;
with ESP32S3.SPI;
procedure SD_Roundtrip is
   C          : Card;
   St         : Status;
   Orig, Back : Block;
begin
   Setup (C, ESP32S3.SPI.SPI2, Sclk => 12, Mosi => 11, Miso => 13, Cs => 10);
   Initialize (C, St);
   Read_Block  (C, 16#2000#, Orig, St);     --  read a scratch sector
   Write_Block (C, 16#2000#, Orig, St);     --  write the same bytes back
   Read_Block  (C, 16#2000#, Back, St);     --  re-read; Back = Orig, no data lost
end SD_Roundtrip;
\end{lstlisting}

\section{SDMMC --- the native SD/MMC host}\index{SDMMC}\index{SDHOST controller}

\texttt{ESP32S3.SDMMC} drives the dedicated SDHOST controller (clock + command +
1- or 4-bit data) in PIO mode --- faster than SPI mode, and with no DMA or
finalization it works under every profile. The API mirrors \texttt{SD\_SPI}.

\begin{lstlisting}
procedure Setup (C : out Card; On : Slot; Clk, Cmd, D0 : Pin_Id;
                 D1, D2, D3 : Optional_Pin := No_Pin;
                 Width : Bus_Width := Width_1; ...);
procedure Initialize (C : in out Card; Result : out Status);
procedure Read_Block  (C : in out Card; LBA : Block_Address; Data : out Block;
                       Result : out Status);
\end{lstlisting}

\textbf{Example --- 4-bit bus, read a sector.}
\begin{lstlisting}
with ESP32S3.SDMMC; use ESP32S3.SDMMC;
procedure SDMMC_Read is
   C   : Card;
   St  : Status;
   Sec : Block;
begin
   Setup (C, On => Slot1, Clk => 14, Cmd => 15, D0 => 2, D1 => 4, D2 => 12,
          D3 => 13, Width => Width_4);
   Initialize (C, St);
   if St = OK then
      Read_Block (C, 0, Sec, St);
   end if;
end SDMMC_Read;
\end{lstlisting}

\section{The ext4 filesystem}\index{ext4 filesystem}\index{lwext4}\index{JBD2 journal}

\texttt{ESP32S3.Ext4} is a pure-Ada ext2/3/4 implementation with a JBD2 journal
(Chapter overview in \texttt{libs/esp32s3\_hal/src/ext4/README.md}). It reads
default \texttt{mkfs.ext4} cards (extents, 64-bit, HTree directories,
\texttt{metadata\_csum} verification), writes ext2/3/non-csum filesystems, and
recovers a dirty journal on mount. It needs the \texttt{embedded} or \texttt{full}
profile.

\subsection{Connecting a block device}\index{block device}

The filesystem talks to a small block-device abstraction (\texttt{ESP32S3.%
Block\_Dev.Device}). An adapter turns an initialised SD card into one with a
single call --- the \texttt{Make} below appears in each complete example that
follows:

\begin{lstlisting}
with ESP32S3.Block_Dev;
with ESP32S3.Block_Dev.SD_SPI_Source;
--  ... given an initialised "Card : aliased ESP32S3.SD_SPI.Card" ...
   Dev : constant ESP32S3.Block_Dev.Device :=
           ESP32S3.Block_Dev.SD_SPI_Source.Make (Card'Access);
\end{lstlisting}

\subsection{Mounting and reading}\index{mount}\index{inode}\index{journal replay}

A \texttt{Mount} is a controlled handle: it flushes and closes on scope exit, and
\texttt{Open} automatically replays a dirty journal on a writable mount. The
complete example brings up the card, wraps it, mounts read-only, then reads a
file:

\begin{lstlisting}
with ESP32S3.SPI;
with ESP32S3.SD_SPI;
with ESP32S3.Block_Dev;
with ESP32S3.Block_Dev.SD_SPI_Source;
with ESP32S3.Ext4;
with ESP32S3.Ext4.FS;
with ESP32S3.Ext4.Inode;
procedure EXT4_Read is
   use type ESP32S3.SD_SPI.Status;
   Card : aliased ESP32S3.SD_SPI.Card;
   St   : ESP32S3.SD_SPI.Status;
begin
   ESP32S3.SD_SPI.Setup (Card, ESP32S3.SPI.SPI2,
                         Sclk => 12, Mosi => 11, Miso => 13, Cs => 10);
   ESP32S3.SD_SPI.Initialize (Card, St);
   if St /= ESP32S3.SD_SPI.OK then
      return;
   end if;
   declare
      Dev  : constant ESP32S3.Block_Dev.Device :=
               ESP32S3.Block_Dev.SD_SPI_Source.Make (Card'Access);
      M    : ESP32S3.Ext4.FS.Mount;
      I    : ESP32S3.Ext4.Inode.Info;
      Buf  : ESP32S3.Ext4.Byte_Array (0 .. 255);
      Last : Natural;
   begin
      M.Open (Dev, Read_Only => True);
      M.Stat (M.Lookup ("/etc/hello.txt"), I);  --  resolve a path -> its inode
      M.Read_File (I, Offset => 0, Into => Buf, Last => Last);
      --  Buf (0 .. Last - 1) holds the first Last bytes of the file
   end;   --  M finalizes: flush + close
end EXT4_Read;
\end{lstlisting}

\textbf{Listing a directory.} \texttt{Iterate} calls a visitor for each entry.
The card bring-up is identical to the example above; the body is:

\begin{lstlisting}
   declare
      Dev  : constant ESP32S3.Block_Dev.Device :=
               ESP32S3.Block_Dev.SD_SPI_Source.Make (Card'Access);
      M    : ESP32S3.Ext4.FS.Mount;
      Root : ESP32S3.Ext4.Inode.Info;
      procedure Show (Name : String;
                      Ino  : ESP32S3.Ext4.Inode_Number;
                      File_Type : ESP32S3.Ext4.U8) is
      begin
         Put_Line ("  " & Name);
      end Show;
   begin
      M.Open (Dev, Read_Only => True);
      M.Stat (M.Lookup ("/"), Root);
      M.Iterate (Root, Show'Access);
   end;
\end{lstlisting}

\subsection{Writing}

On a non-\texttt{metadata\_csum} filesystem (ext2/3, or
\texttt{mke2fs -O \textasciicircum metadata\_csum}) the filesystem can be modified.
The full POSIX-like set is available: \texttt{Create\_File}, \texttt{Write\_File},
\texttt{Truncate}, \texttt{Mkdir}, \texttt{Rmdir}, \texttt{Unlink},
\texttt{Rename} and hard \texttt{Link}.

\begin{lstlisting}
   declare
      Dev : constant ESP32S3.Block_Dev.Device :=
              ESP32S3.Block_Dev.SD_SPI_Source.Make (Card'Access);
      M   : ESP32S3.Ext4.FS.Mount;
      N   : ESP32S3.Ext4.Inode_Number;
   begin
      M.Open (Dev, Read_Only => False);
      N := M.Create_File ("/", "log.txt");
      M.Write_File (N, (Character'Pos ('h'), Character'Pos ('i'), 10));
      M.Mkdir ("/", "data");
      M.Close;          --  flush + close (or let M finalize)
   end;
\end{lstlisting}

\subsection{Streaming writes and file-size limits}\index{streaming write}\index{Append}\index{block map}\index{single-indirect block}\index{double-indirect block}

\texttt{Write\_File} sets a file's whole contents from one in-memory buffer ---
fine for a small file, but you cannot hand an embedded part a gigabyte-sized
array. \texttt{Append} grows a file at its end one chunk at a time, so a large
file is built from small writes with no buffer to hold it all:

\begin{lstlisting}
   N := M.Create_File ("/", "big.log");
   for I in 1 .. 1000 loop          --  e.g. 1000 records, streamed
      M.Append (N, Chunk);          --  grows the file; no whole-file buffer
   end loop;
   M.Commit;
\end{lstlisting}

What bounds a single file is the \emph{block map}. The writer allocates classic
indirect blocks --- 12 direct, then single-indirect, then double-indirect --- so
on a 4\,KiB-block volume one file reaches about 4\,MiB through the single-indirect
tier and about 4\,GiB through the double-indirect tier; the matching free and
truncate paths walk the same tiers, so a file that can be written can always be
deleted and shrunk. The \emph{read} path goes further --- it follows the full
classic map (through triple-indirect) and ext4 extent trees --- so a
multi-gigabyte file written by a host is readable on-device, bounded only by the
volume's free space.

\subsection{Crash-safe transactions}\index{crash-safe transactions}\index{journaling}\index{commit barrier}

For atomicity, group an operation behind \texttt{Commit}: the dirty metadata and
the superblock are written to the journal (durably), the RECOVER flag is set ---
the commit barrier --- then the metadata is checkpointed to its final locations
and the flag cleared. A power loss after the barrier is recovered forward by the
next \texttt{Open} (or by any other ext4 implementation, including the Linux
kernel); a loss before it has no effect.

\begin{lstlisting}
   --  Payload : constant ESP32S3.Ext4.Byte_Array := (16#41#, 16#42#, 16#43#);
   N := M.Create_File ("/", "important.txt");
   M.Write_File (N, Payload);
   M.Commit;         --  one atomic, journaled, crash-safe transaction
\end{lstlisting}

A volume formatted \emph{without} a journal (\texttt{mke2fs -O \textasciicircum
has\_journal}) has no log to write through, so \texttt{Commit} instead flushes the
dirty metadata and superblock straight to their final locations. The filesystem
detects which kind it mounted and routes \texttt{Commit} automatically --- a
no-journal volume trades the crash-safety barrier for smaller size and fewer
writes, the right choice on a small flash chip.\index{no-journal volume}

\subsection{Concurrency: one owner per mount}\index{block cache}\index{single-owner mount}

A \texttt{Mount} is a \emph{single-owner} handle, not a shared service. Every
operation takes it \texttt{in out} and mutates shared state, and the filesystem
holds no internal lock or protected object: drive one mount from one task, and let
its calls run to completion in sequence.

That rule is stricter than it first looks, because even reads have side effects.
Underneath the mount sits a write-back LRU block cache, and \texttt{Read\_File}
fetches-on-miss and reorders it --- so two tasks ``only reading'' the same mount
still race on the cache. Writes are stronger still: \texttt{Create\_File},
\texttt{Write\_File}, \texttt{Mkdir} and the rest accumulate a dirty-metadata
write-set that \texttt{Commit} flushes as \emph{one} journaled transaction (the
crash-safe commit above). Interleaving two tasks' writes scrambles that write-set
and breaks the crash-safety guarantee, which assumes a single writer building one
coherent transaction. The block drivers \emph{beneath} the filesystem
(\texttt{SD\_SPI}, \texttt{SDMMC}) are themselves task-safe, but that only
serialises the \emph{bus}; it does nothing for the filesystem state above it.

If more than one task must reach the filesystem, give the mount a single owner and
funnel every request through it: a dedicated task that owns the \texttt{Mount} and
performs the I/O, with other tasks handing it work through a \texttt{protected}
request queue (Ravenscar-compatible) or, on the \texttt{full} profile, a rendezvous
entry. The I/O itself must stay in the owning task --- never inside a protected
action, because the SD drivers can suspend and a protected operation must not block.

\begin{lstlisting}
--  One task owns the Mount and is the only code that touches it; other tasks
--  hand it work (a protected request queue under Ravenscar, or a rendezvous
--  entry on the full profile).  The blocking I/O stays in this task.
task body File_Server is
   M : ESP32S3.Ext4.FS.Mount;          --  the single owner
begin
   M.Open (Dev, Read_Only => False);
   loop
      --  take the next request, then service it on the one mount:
      --     M.Read_File (...)              --  a read request
      --     M.Write_File (...);  M.Commit; --  a write request, committed
      null;
   end loop;
end File_Server;
\end{lstlisting}

\subsection{Errors}

Filesystem operations raise: the standard \texttt{Ada.IO\_Exceptions} names
(\texttt{Name\_Error} for a missing path, \texttt{Use\_Error}, \texttt{Device\_%
Error}, \texttt{End\_Error}) plus filesystem-specific \texttt{Unsupported\_%
Feature}, \texttt{Bad\_Checksum}, \texttt{Corrupt}, \texttt{No\_Space},
\texttt{Not\_Empty} and \texttt{Read\_Only}. Wrap calls in a handler so a bad card
or a missing file never escapes a task:

\begin{lstlisting}
begin
   M.Stat (M.Lookup ("/maybe/missing"), I);
exception
   when ESP32S3.Ext4.Name_Error => Put_Line ("not found");
   when ESP32S3.Ext4.Bad_Checksum => Put_Line ("corrupt metadata");
end;
\end{lstlisting}

\subsection{Free-count integrity: the phantom-free tripwire}\index{phantom-free tripwire}\index{free-count drift}

A subtle class of corruption is the \emph{free-count drift}: the superblock and
block-group descriptors cache how many blocks and inodes are free, and if a
\texttt{Free} adjusts that cache for a bit that was \emph{already} clear --- a
double-free, or a stale/incoherent bitmap read off a flaky SD card --- the cached
count creeps above what the bitmap actually says. \texttt{e2fsck} then reports a
Pass-5 count mismatch on an otherwise-sound filesystem.

The fix makes \texttt{Free} \emph{idempotent} by construction. \texttt{Clear\_Bit}
now reports whether the bit was actually set, and the group/superblock counts are
bumped \emph{only} when it was --- so a double or phantom free is a no-op rather
than a drift, and the count cannot diverge from the bitmap.

\begin{lstlisting}
--  Bitmap.Clear_Bit returns False if the bit was already clear:
if Bitmap.Clear_Bit (Group_BM, Index) then
   Dec_Used (Group);          --  bit really was set: adjust the counts
else
   Phantom_Frees := Phantom_Frees + 1;   --  already free: no drift, but flag it
end if;
\end{lstlisting}

Crucially the already-clear path is \emph{not} silent --- that would mask a genuine
double-free bug. A \texttt{Phantom\_Free\_Count} tripwire counts the events (a plain
counter, not a \texttt{pragma Assert}, because the target also builds with
\texttt{-gnata} and an assert would abort the filesystem operation). The host test
harness asserts the count is zero after every scenario, and the on-target
\texttt{esp32s3\_ext4\_write} example reports it as a battery check
(\texttt{[PASS] no phantom frees}; a flaky card flips it to \texttt{[FAIL]}).

\begin{lstlisting}
Bitmap.Reset_Phantom_Free_Count;            --  arm the tripwire
--  ... exercise allocate / free / truncate / unlink ...
Check ("no phantom frees (count consistent with bitmap)",
       Bitmap.Phantom_Free_Count = 0);
\end{lstlisting}

The original drift was ultimately traced to a \emph{worn} SD card: a brand-new card
runs the whole battery \texttt{e2fsck}-clean with zero phantom frees. The tripwire
is what turned an intermittent, card-dependent count mismatch into a single
yes/no signal --- surfacing the fault instead of hiding it.

\subsection{How it is validated}

Because the filesystem is pure logic over the block interface, it is developed and
tested host-native (x86) against real \texttt{mke2fs} images, with the Linux
kernel and \texttt{e2fsck} as the oracle: every read is byte-compared to the
source file, every mutation is checked with \texttt{e2fsck -fn}, and journal
replay and commit are cross-checked against the kernel's own recovery. The same
code cross-compiles for the ESP32-S3 and runs over the SD drivers above. See
\texttt{tests/ext4\_host/} and \texttt{libs/esp32s3\_hal/src/ext4/REFERENCES.md}.

\section{Putting ext4 on raw SPI NOR flash}\index{SPI NOR flash}\index{wear leveling}\index{flash translation layer}

The same filesystem runs on a bare SPI NOR flash chip, no SD card in sight. Four
pure-Ada layers stack up, each a thin adapter over the one below:

\begin{center}
\texttt{ESP32S3.Ext4} $\rightarrow$ \texttt{Block\_Dev.WL} $\rightarrow$
\texttt{Block\_Dev.W25Q\_Source} $\rightarrow$ \texttt{ESP32S3.W25Q}
\end{center}

\noindent The flash driver speaks to the chip; \texttt{W25Q\_Source} presents it
as a 512-byte block device; \texttt{WL} is a wear-leveling flash translation
layer that spreads writes so no sector wears out first; and \texttt{Ext4} mounts
the result. The reference part is a Winbond \texttt{W25Q256FV} (32\,MB) that shares
the SPI2 bus with the W5500 Ethernet chip.

\subsection{The W25Q flash driver}\index{W25Q}\index{JEDEC ID}\index{4-byte addressing}\index{chip-select callback}

\texttt{ESP32S3.W25Q} reads the JEDEC id, then programs, erases and reads in
\emph{4-byte address mode} (a part larger than 16\,MB needs four address bytes).
Two design points stand out. The chip size is \emph{detected}, not hard-coded: the
JEDEC density byte is a power of two, so \texttt{Capacity\_Bytes} turns
\texttt{EF\,40\,19} into 32\,MB, and because each layer above sizes itself from the
one below, fitting a different W25Q part (8/16/32/64\,MB) needs no code change.
And the chip-select is \emph{application-driven}: sharing a bus with another
device, the flash brings its own active-low select on a GPIO through a callback
(\texttt{GPIO\_Select} over IO21 here), rather than the SPI host's single hardware
\texttt{CS0}.

\begin{lstlisting}
with ESP32S3.W25Q; use ESP32S3.W25Q;
--  Select_Pad : aliased Pin_Cell := (Pin => 21);    --  active-low CS on IO21
--  Flash : ESP32S3.W25Q.Flash :=
--    (Host => ESP32S3.SPI.SPI2, CS => GPIO_Select'Access,
--     Ctx => Select_Pad'Address);
   ID      : JEDEC_ID;
   Mode_OK : Boolean;
begin
   Read_Identification (Flash, ID);          --  EF 40 19 on a W25Q256FV
   Initialize (Flash, Mode_OK);              --  enter 4-byte address mode
   if ID.Manufacturer = 16#EF# and then Capacity_Bytes (ID) /= 0 then
      null;                                  --  Capacity_Bytes (ID) = 32 MB
   end if;
\end{lstlisting}

\subsection{Block device and wear leveling}\index{W25Q\_Source}\index{block-erase}\index{ping-pong configuration}

NOR flash must be \emph{erased} (to all-ones, in 4\,KiB sectors) before it can be
re-programmed, so \texttt{W25Q\_Source} hides an erase-aware, write-through
512-byte sector interface over the chip; an optional block-erase capability lets
the layer above clear a whole 4\,KiB block in one operation.

\texttt{Block\_Dev.WL} is the wear-leveling FTL. It keeps an $O(1)$ state --- a
single move counter --- and maps a logical 4\,KiB block to a physical one by a
shifting offset, with one always-free ``hole''. Every few writes it advances the
hole by copying one block, so over time every logical block visits every physical
block and a hot block (an ext4 bitmap, say) never wears out one sector. The state
is written to two configuration blocks \emph{ping-pong} (alternating, each with a
higher sequence number and a CRC), so a power loss mid-update always leaves a
consistent earlier state; \texttt{Mount} picks the highest valid sequence. The
whole stack --- erase-aware writes, the wear-level remap, the auto-detected size
--- comes together in four calls:

\begin{lstlisting}
   BDW.Configure (Raw, Flash => Flash);      --  auto-size from the JEDEC id
   WL.Attach (Vol, BDW.Make (Raw'Access));   --  wear-leveling volume over it
   WL.Mount  (Vol, Formatted);               --  recover the saved wear state
   Dev := WL.Make (Vol'Access);              --  a 512-byte Block_Dev for ext4
\end{lstlisting}

\subsection{Formatting on-device: \texttt{Ext4.Mkfs}}\index{mkfs}\index{Ext4.Mkfs}\index{on-device format}

The external flash is not on the host flasher's path, and the filesystem reads and
writes but does not format --- so to start from a blank chip the board makes the
filesystem itself. \texttt{ESP32S3.Ext4.Mkfs} is a minimal pure-Ada
\texttt{mkfs.ext4}: it writes a single-block-group ext4 (4\,KiB blocks, classic
block-mapped inodes, a root directory and \texttt{lost+found}), with an optional
4\,MiB JBD2 journal (\texttt{Journal => True}). It is the inverse of the read path,
validated against the host's \texttt{e2fsck}. From a blank chip to a mounted,
written filesystem is then:

\begin{lstlisting}
   --  ... build Dev as above, but WL.Format (Vol) for a fresh volume ...
   ESP32S3.Ext4.Mkfs.Format (Dev, Volume_Label => "FLASH", Journal => True);
   M.Open (Dev, Read_Only => False);
   N := M.Create_File ("/", "boot.txt");
   M.Write_File (N, (Character'Pos ('h'), Character'Pos ('i'), 10));
   M.Commit;                                 --  through the journal just created
\end{lstlisting}

\noindent The on-target examples \texttt{esp32s3\_w25q}, \texttt{esp32s3\_wl} and
\texttt{esp32s3\_ext4\_mkfs} run this stack end to end on the hardware.

\vfill
\noindent\rule{\linewidth}{0.4pt}\\[4pt]
\textit{The whole stack --- runtime, peripheral HAL, and filesystem --- is one
coherent body of Ada: strong types from the register up, the same RAII and
task-safety patterns in every driver, and a filesystem validated against the
Linux kernel, all running on both cores of the ESP32-S3 on its own bare-metal
runtime.}
